Exponential Random Graph Models with Big Networks: Maximum Pseudolikelihood Estimation and the Parametric Bootstrap

### 2017-07-26 TeX/conference_paper.tex
revising intro

\section{The Exponential Random Graph Model}
\noindent The \textit{exponential random graph model} (ERGM) is a probability model for directed or undirected binary networks. This means neither the weighting nor the temporal change of ties is considered in the model.
In literature, ERGMs are sometimes also referred to as \textit{p-star} or $p^*$ models (see Wassermann and Pattison \cite{Wasserman.1996}, Robins et al. \cite{robins.pattison.2007}). In the following, we will introduce the ERGM for directed networks; however, undirected networks can be introduced very similar.\\[0.3cm] 
The ERGM takes the adjacency matrix of an observed network $G^{obs}$ as the manifestation of a matrix-like random variable $Y$. This means that a network of $N$ nodes can be defined as a adjacency matrix $G=(g_{ij})\in \mathbb{R}^{N \times N}$, where $g_{ij} \in \{0,1\}$ for all $i,j \in \{1,\dots , N\}$. $g_{ij}=1$ means that there is an edge between actors $i$ and $j$, while $g_{ij}=0$ indicates that these actors are not directly connected. Since the model does not consider loops, one has $g_{ii}=0$ for all $i \in \{1,\dots , N\}$. Furthermore, define
$$ \mathcal{G}(N) := \left\{ G \in \mathbb{R}^{(N \times N)}: g_{ij} \in \{0,1\},~g_{ii}=0\right\}$$
as the set of all possible networks on $N$ nodes without loops. Note that the cardinality of set $\mathcal{G}(N)$ is increasing exponentially for every newly included actor, which results in $2^{N(N-1)/2}$ total elements. Therefore, for an already small number of actors the cardinality of $\mathcal{G}(N)$ turns out to be an astronomically large number. For this reason calculating the MLE is either extremely time-consuming or with today's technology not achievable. As a consequence, literature usually makes use of Markov Chain Monte Carlo (MCMC) methods as we will intriduce in the next chapter.\\  
With the definition of $\mathcal{G}(N)$ we define 
$$Y: \Omega \to \mathcal{G}(N)~~,~~\omega \mapsto (Y_{ij}(\omega))_{i,j=1,\dots,N}$$
as a matrix-like random variable. As the probability function from $Y$ to $\mathcal{G}(N)$ we define the ERGM as
\begin{equation}
\mathbb{P}_{\theta}(Y=G)=\dfrac{\exp(\theta^T \cdot \Gamma(G))}{\sum_{G^* \in \mathcal{G}(N)} \exp(\theta^T \cdot \Gamma(G^*))} 
\label{ERGM}
\end{equation}
%
where $\theta \in \mathbb{R}^q$ is a $q-$dimensional vector of parameters, $\Gamma:\mathcal{G}(N) \to \mathbb{R}^q~,~G \mapsto (\Gamma_1(G),\dots,\Gamma_q(G))^T$ is a $q$-dimensional function of different network statistics and $c(\theta):= \sum_{G^* \in \mathcal{G}(N)} \exp(\theta^T \cdot \Gamma(G^*))$ is a normalization constant which ensures that (\ref{ERGM}) defines a probability function on $\mathcal{G}$.
%
As already mentioned, a specific network $G$ can be considered as a manifestation of a matrix-like random variable, whose probability of occurrence can be modeled with equation (\ref{ERGM}).
A key role when modeling an ERGM is played by the function $\Gamma(\cdot)$. The decision about which network statistics are incorporated into the model affects the model significantly (see Handcock \cite{Handcock.2003}).

>>>>

\section{The Exponential Random Graph Model}
\noindent The \textit{exponential random graph model} (ERGM) is a probabilistic model for networks \cite{chatterjee2013estimating,Wasserman.1996,robins.pattison.2007}. They can be used for link prediction \cite{lu2010supervised}, simulating network adjacency matrices \cite{hackney2006agent}, and testing theories regarding the processes underlying tie formation \cite{goodreau2009birds}. The ERGM was first introduced by Holland and Leinhardt (1981) \cite{holland1981exponential}. However, due to the intractable normalizing constant in the likelihood function of the ERGM, it did not see widespread and complete use until the 2000s, following the development of algorithms and software for efficient simulation-based methods for working with ERGM \cite{snijders2002markov}. Training ERGM using simulation-based methods is computationally expensive, and can still be prohibitively burdensome with data on big networks. Approximate methods of estimation, which are much more feasible with large networks, have existed for some time, but these methods perform poorly when it comes to characterizing the uncertainty in parameter estimates, which is necessary when assessing risk in predictions or simulation, or in hypothesis testing.

The ERGM takes the adjacency matrix of an observed network $G^{obs}$ as the manifestation of a matrix-like random variable $Y$. This means that a network of $N$ nodes can be defined as a adjacency matrix $G=(g_{ij})\in \mathbb{R}^{N \times N}$, where $g_{ij} \in \{0,1\}$ for all $i,j \in \{1,\dots , N\}$. $g_{ij}=1$ means that there is an edge between actors $i$ and $j$, while $g_{ij}=0$ indicates that these actors are not directly connected. Since the model does not consider loops, one has $g_{ii}=0$ for all $i \in \{1,\dots , N\}$. Furthermore, define
$$ \mathcal{G}(N) := \left\{ G \in \mathbb{R}^{(N \times N)}: g_{ij} \in \{0,1\},~g_{ii}=0\right\}$$
as the set of all possible networks on $N$ nodes without loops. Note that the cardinality of set $\mathcal{G}(N)$ is increasing exponentially for every newly included actor, which results in $2^{N(N-1)/2}$ total elements. Therefore, for an already small number of actors the cardinality of $\mathcal{G}(N)$ turns out to be an astronomically large number. For this reason calculating the MLE is either extremely time-consuming or with today's technology not achievable. As a consequence, literature usually makes use of Markov Chain Monte Carlo (MCMC) methods as we will intriduce in the next chapter.\\  
With the definition of $\mathcal{G}(N)$ we define 
$$Y: \Omega \to \mathcal{G}(N)~~,~~\omega \mapsto (Y_{ij}(\omega))_{i,j=1,\dots,N}$$
as a matrix-like random variable. As the probability function from $Y$ to $\mathcal{G}(N)$ we define the ERGM as
\begin{equation}
\mathbb{P}_{\theta}(Y=G)=\dfrac{\exp(\theta^T \cdot \Gamma(G))}{\sum_{G^* \in \mathcal{G}(N)} \exp(\theta^T \cdot \Gamma(G^*))} 
\label{ERGM}
\end{equation}
%
where $\theta \in \mathbb{R}^q$ is a $q-$dimensional vector of parameters, $\Gamma:\mathcal{G}(N) \to \mathbb{R}^q~,~G \mapsto (\Gamma_1(G),\dots,\Gamma_q(G))^T$ is a $q$-dimensional function of different network statistics and $c(\theta):= \sum_{G^* \in \mathcal{G}(N)} \exp(\theta^T \cdot \Gamma(G^*))$ is a normalization constant which ensures that (\ref{ERGM}) defines a probability function on $\mathcal{G}$.
%
As already mentioned, a specific network $G$ can be considered as a manifestation of a matrix-like random variable, whose probability of occurrence can be modeled with equation (\ref{ERGM}).
A key role when modeling an ERGM is played by the function $\Gamma(\cdot)$. The decision about which network statistics are incorporated into the model affects the model significantly (see Handcock \cite{Handcock.2003}).
